---
title: "HiLDA: a package for testing the burdens of mutational signatures"
author: 
- name: Zhi Yang
  affiliation: Department of Preventive Medicine, 
    University of Southern California, Los Angeles, USA 
  email: zyang895@gmail.com
date: "`r Sys.Date()`"
vignette: |
  %\VignetteIndexEntry{An introduction to HiLDA}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
output:
  BiocStyle::html_document:
    toc_float: true
  BiocStyle::pdf_document: default
abstract: | 
  Instructions on using _HiLDA_ on testing the burdens of mutational signatures. 

---

```{r style, echo = FALSE, results = 'asis'}
library(BiocStyle)
```

# Introduction

The R package **HiLDA** is developed under the Bayesian framework to allow 
statisticaly testing whether there is a change in the mutation burdnes of 
mutation sigantures between two groups. The mutation signature is defined based 
on the independent model proposed by Shiraishi's et al. 

# Paper

- Shiraishi et al. A simple model-based approach to inferring and visualizing 
cancer mutation signatures, bioRxiv, doi: 
[http://dx.doi.org/10.1101/019901](http://dx.doi.org/10.1101/019901).

- **Zhi Yang**, Priyatama Pandey, Darryl Shibata, David V. Conti, Paul Marjoram, 
Kimberly D. Siegmund. HiLDA: a statistical approach to investigate differences 
in mutational signatures, bioRxiv, doi: https://doi.org/10.1101/577452 


# Input data

`HiLDA` is a package built on some basic functions from `pmsignature` including 
how to read the input data. Here is an example from `pmsignature` on the input 
data, *mutation features* are elements used for categorizing mutations such as: 
  
* 6 substitutions (C>A, C>G, C>T, T>A, T>C and T>G)
* 2 flanking bases (A, C, G and T)
* transcription direction.

## Mutation Position Format

sample1 chr1  100	A	C	
sample1	chr1	200	A	T	
sample1	chr2	100	G	T	
sample2	chr1	300	T	C	
sample3	chr3	400	T	C	
  
* The 1st column shows the name of samples 
* The 2nd column shows the name of chromosome 
* The 3rd column shows the coordinate in the chromosome
* The 4th column shows the reference base (A, C, G, or T).
* The 5th colum shows the alternate base (A, C, G, or T).


# Workflow 
## Installation 
### Install the packages
First, a few R packages such as `pmsignature`, `gtools`, `R2jags` have to be 
installed prior to using `HiLDA`. Currently, the easiest way for installing 
pmsignature is to use the package devtools. 

```{r eval=FALSE}
devtools::install_github("USCbiostats/HiLDA")
```

### Install `JAGS`
Download and install JAGS by following the instructions from 
http://mcmc-jags.sourceforge.net/

## Get input data
Here, *inputFile* is the path for the input file. *numBases* is the number of 
flanking bases to consider including the central base (if you want to consider 
two 5' and 3' bases, then set 5). Also, you can add transcription direction 
information using *trDir*. *numSig* sets the number of mutation signatures 
estimated from the input data.  
```{r}
library(HiLDA)
inputFile <- system.file("extdata/esophageal.mp.txt.gz", package="HiLDA")
G <- hildaReadMPFile(inputFile, numBases=5, trDir=TRUE)
```

Loading a simulated dataset G with 10 mutational catalogs. 
```{r}
load(system.file("extdata/sample.rdata", package = "HiLDA"))
```

## Run tests from `HiLDA`
### Perform the global test and the local test 
```{r}
hildaGlobal <- hildaGlobalTest(inputG=G, numSig=3, refGroup=1:4, nIter=1000)
hildaLocal <- hildaLocalTest(inputG=G, numSig=3, refGroup=1:4, nIter=1000)
```

### Perform the global test and the local test with initial values
```{r}
load(system.file("extdata/sampleParam.rdata", package="HiLDA"))

hildaGlobal <- hildaGlobalTest(inputG=G, numSig=3, useInits = Param,
                               refGroup=1:4, nIter=1000)
hildaLocal <- hildaLocalTest(inputG=G, numSig=3, useInits = Param,
                             refGroup=1:4, nIter=1000)
```

###  Assess Convergence of MCMC chains
```{r}
hildaRhat(hildaGlobal)
hildaRhat(hildaLocal)
```

## Get signatures from *pmsignature*
```{r}
load(system.file("extdata/sampleParam.rdata", package="HiLDA"))
```

## Visualize the mutation signatures from pmsignature
```{r}
pmPlots <- pmBarplot(G, Param, refGroup=1:4, sigOrder=c(1,3,2))
cowplot::plot_grid(pmPlots$sigPlot, pmPlots$propPlot, rel_widths = c(1,3))
```

## Visualize the mutation signatures from HiLDA
```{r}
hildaPlots <- hildaBarplot(G, hildaLocal, refGroup=1:4, sigOrder=c(1,3,2))
cowplot::plot_grid(pmPlots$sigPlot, pmPlots$propPlot, rel_widths = c(1,3))
```

## Output the posterior distribution of the mean difference in mutational 
## exposures
```{r}
hildaDiffPlots <- hildaDiffPlot(G, hildaLocal, sigOrder=c(1,3,2))
cowplot::plot_grid(hildaDiffPlots$sigPlot, hildaDiffPlots$diffPlot, 
                   rel_widths = c(1,3))
```


